Functional Pearl \\ Programming with Negative and Fractional Types

### 2012-02-19 dual.tex
reworded fractional discussion

Consider the following two ways of purchasing an item that costs \$20.00:
\begin{enumerate}
\item You give the seller a \$20.00 bill.
\item You use a credit card to pay the seller.
\end{enumerate}
In both cases, the seller receives the money immediately but there is a
subtle difference. In the first transaction, the money received by the seller
corresponds to a value that has been produced earlier in the computation. In
the second transaction, the money may or may not exist yet: computationally,
a \emph{debt} is generated to compensate for the money received by the seller
and this debt travels ``backwards'' towards the bank and is (hopefully)
reconciled there.

@@

For this reason, our treatment of negative values in the context of
{{langRev}} will be particularly simple. We will have a type $0$ and an
isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
allows us to create a value and a corresponding debt out of nothing. Both the
value and its negative counterpart can flow anywhere in the system. Because
information is preserved, a closed program (which does not have any
``dangling'' negative value) will eventually have to match the negative value
with some corresponding value, effectively using the isomorphism in the
right-to-left direction. In more detail, we can model the credit card
transaction above using the following program (shown diagrammatically):

@@

The type $a/b$ is prominent in the Lambek-Grishin calculus which is
extensively used in computational linguistics. In the common interpretation,
a value of type $a/b$ is a value of some type $c$ such that when put in a
context of type $b$, the result is a value of type $a$. For example, assuming
contexts are represented as functions, a value of type $\texttt{bool} /
(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
type $\texttt{bool}$. More generally, one can think of a value of type $a/b$
as a value that imposes a constraint on its context of use: it is a value
that requires its context to be of type $b$ and only if that condition is
satisfied, can the value be considered as a value of type $a$. In the
simplest case, the type $1/b$ just constraints the context to be of type $b$.

By allowing the fractional types to be first-class, we can separate the
generation of constraints from their use. For example, consider a variation
on the example in the previous where the parent gives a child the \$20.00
bill but requires that the child only spends the money on buying books. This
can achieved by producing a constraint that can only be resolved if the child
indeed buys a book. Technically, this is achieved by adding a new type
isomorphism {{ 1 <-> a * (1/a) }} which when used in the left-to-right
direction allows the creation of a value and a corresponding constraint. Both
the value and the constraint can flow in arbitrary ways during the
computation but eventually in a closed program with no ``dangling''
constraints, the isomorphism should be used in the right-to-left direction to
resolve the constraints.

NEED to work out the example in detail (or perhaps another example that's
better. Not sure if this one makes sense because we have to create the book
and the context at once.???

>>>>

Consider the following two ways of purchasing an item that costs \$20.00:
\begin{enumerate}
\item You give the seller a \$20.00 bill.
\item You use a credit card to pay the seller.
\end{enumerate}
In both cases, the seller receives the money immediately but there is a
subtle difference. In the first transaction, the money received by the seller
corresponds to a value that has been produced earlier in the computation. In
the second transaction, the money may or may not exist yet: computationally,
a \emph{debt} is generated to compensate for the money received by the seller
and this debt travels ``backwards'' towards the bank where it is (hopefully)
reconciled.

@@

For this reason, our treatment of negative values in the context of
{{langRev}} is particularly simple. We will have a type $0$ and an
isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
allows us to create a value and a corresponding debt out of nothing. Both the
value and its negative counterpart can flow anywhere in the system. Because
information is preserved, a closed program (which does not have any
``dangling'' negative values) will eventually have to match the negative
value with some corresponding value, effectively using the isomorphism in the
right-to-left direction. In more detail, we can model the credit card
transaction above using the following program (shown diagrammatically):

@@

The type $a/b$ is prominent in the Lambek-Grishin calculus which is
extensively used in computational linguistics. In the common interpretation,
a value of type $a/b$ is a value of some type $c$ such that when put in a
context of type $b$, the result is a value of type $a$. For example, assuming
contexts are represented as functions, a value of type $\texttt{bool} /
(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
type $\texttt{bool}$.

In our case, the situation is simpler. We have an isomorphism 
{{ 1 <-> a * (1/a) }} which when used in the left-to-right direction allows the
creation of a value and a corresponding constraint on the context. 
Both the value and the constraint can flow in arbitrary ways during the computation 
but eventually in a closed program with no ``dangling'' constraints, the 
isomorphism should be used in the right-to-left direction to resolve the constraints. 
To understand the computational interpretation, consider the following example.

The goal is to produce a value of type $(\texttt{bool} \times
\texttt{int})$. One part of the program, $e_1$, knows how to produce the
value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
to produce the value of type $\texttt{bool}$ (say {{true}}). Computationally,
we model this as follows. The first subcomputation $e_1$ produces {{((),3)}}
and uses the isomorphism to convert {{()}} to {{(alpha,1/alpha)}} where
{{alpha}} is a yet-unknown boolean value. Reshuffling the components, $e_1$
produces {{((alpha,3), 1/alpha)}}. Similarly, $e_2$ can independently produce
{{((true,beta),1/beta)}} where {{beta}} is a yet-unknown integer value.
When the two values meet, we can group the components as follows:
\[
{{(alpha,beta)}}  {{(3,1/beta)}}  {{(true,1/alpha)}}
\]
Using the isomorphism in the right-to-left direction on the last two pairs,
forces {{alpha}} to be resolved to {{true}} and forces {{beta}} to be
resolved to {{3}}. Both of these pairs then become {{()}} and be absorbed. We
are left with the pair {{(true,3)}} as desired.

To summarize, one can think of a value of type $a/b$ as a value that imposes
a constraint on its context of use: it is a value that requires its context
to be of type $b$ and only if that condition is satisfied, can the value be
considered as a value of type $a$. In the simplest case, the type $1/b$ just
constraints the context to be of type $b$. By allowing the fractional types
to be first-class, we can separate the generation of constraints from their
use. Both the value and the constraint can flow in arbitrary ways during the
computation but eventually in a closed program with no ``dangling''
constraints, the isomorphism should be used in the right-to-left direction to
resolve the constraints.

NEED to work out the example in detail (or perhaps another example that's
better.
